Non-convergence in iterative imputation

### 2025-04-17 Manuscript/manuscript.qmd
merge methods sections

---
title: "Non-convergence in iterative imputation"
author: "H. I. Oberman"
abstract: "Iterative imputation has become the de facto standard to accommodate for the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms. We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached. In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value."
# keywords: "missing data, iterative imputation, non-convergence, MICE"
# bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
format: pdf
bibliography: references.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
load("../results/out.RData")
source("../R/evaluate_estimates.R")
performance <- evaluate_est(out)
# performance[performance$term == "b0", "term"] <- "Y"
results <- # simulation results split by condition and regression coefficient
performance |> 
  group_by(method, mech, prop, .it, term) |> 
  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) |>
  mutate(mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
         prop = factor(prop, levels = c(0.25, 0.5, 0.75), ordered = TRUE),
         term = factor(term, levels = c("Y", "X1", "X2", "X3"), ordered = TRUE))
```

@@

We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; R Core Team 2024). We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI].

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 2000$). To assess non-convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 100$). We differentiate between nine different missingness scenarios that are defined in the data generating mechanism.

@@

We generate a complete set of $n_{\mathrm{obs}} = 200$ cases in each simulation repetition. This set consists of three multivariately normal random variables, such that

$$
\begin{pmatrix}
X_1\\
X_2\\
X_3
\end{pmatrix}
\sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
0\\
0\\
0
\end{pmatrix},
\begin{pmatrix}
1     &       &  \\
0.5   & 1     &  \\
0.5  & 0.5   & 1
\end{pmatrix}
\end{bmatrix},
$$ 
and an outcome variable $Y$ as a linear combination of the three predictors. For each unit $i = 1, 2,..., n_\mathrm{obs}$,

$$
Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 1)$. This results in a complete dataset all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$) in each simulation repetition. Multivariate normal data are generated using the `mvtnorm` package [@mvtnorm].

Subsequently, the complete set is 'amputed' (i.e., made incomplete) according to six missingness conditions. We use a $2 \times 3$ factorial design with three missingness mechanisms and three proportions of incomplete cases, see Table XYZ. Missingness mechanisms refer to the probability of being missing for any given entry in a dataset. There are three distinct types of mechanisms, as defined by Rubin (1976). Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), may depend on observed information (MAR; Missing At Random), or may depend on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random). [TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018; TODO: add `ampute()`].

@@

We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice]. On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest. We predict outcome variable $Y$ from the other variables,

@@

where $\hat{Y}$ is the predicted value of the outcome. Our estimands are the regression coefficients $\boldsymbol{\beta}$, including the intercept $\beta_0$.

@@

$$
\begin{aligned}
\hline
&\mathrm{Non-convergence \ diagnostics} & & & & \\
\hline
&\begin{matrix}
\widehat{R} \\
\mathrm{AC } 
\end{matrix}
\quad
&\times
\quad
&\begin{matrix}
\mathrm{chain \ means} \\
\mathrm{chain \ variances} 
\end{matrix}
\quad
&\times
\quad
&\begin{matrix}
Y\\
X_1\\
X_2\\
X_3
\end{matrix} \\
&\mathrm{identifier} &&\mathrm{parameter} &&\mathrm{variable} \\
\hline
\end{aligned}
$$

As recommended by van Buuren (-@buur18 $\S$ 2.5.2), our performance measures are bias, confidence interval width, and coverage rate of the estimands. These measures are averaged across the simulation repetitions.
<!-- [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!]. -->

@@

```{r ac, fig.ncol=2}
results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = ac_mean, color = mech, linetype = mech)) +
  geom_hline(yintercept = 0, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()

results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = ac_sd, color = mech, linetype = mech)) +
  geom_hline(yintercept = 0, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()
```

@@

```{r psrf, fig.ncol=2}
results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = psrf_mean, color = mech, linetype = mech)) +
  geom_hline(yintercept = 1, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()

results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = psrf_sd, color = mech, linetype = mech)) +
  geom_hline(yintercept = 1, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()
```

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $y_{\mathrm mis}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot y_{\mathrm imp, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $y_{\mathrm obs}$ to create $m$ completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\mathrm obs}$; the missing part of $y$ is referred to as $y_{\mathrm mis}$. For each datapoint in $y_{\mathrm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$). The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.

@@

To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays two scenarios from the example. The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations). The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm imp, \ell}$.

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm imp, \ell}$. This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$. $\bar{Q}$ is then a  biased, invalid estimate of $Q$. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.

@@

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation. The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}. By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.

@@

To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'. Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\mathrm inc}$. Early stopping is defined by the number of iterations in the imputation algorithm, $T$.

The missingness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\mathrm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. A higher $p_{\mathrm inc}$ yields slower convergence---unless there is so little information in $y_{\mathrm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that missingness conditions with higher $p_{\mathrm inc}$ will result in more signs of non-convergence.

@@

1. We expect that simulation conditions with a high proportion of incomplete cases $p_{\mathrm inc}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.

@@

$$
Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 1)$. This results in a dataset $y$, where $y_{\mathrm mis} = \emptyset$. In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).

@@

In each simulation repetition, we *ampute* the complete dataset conform six missingness conditions: two missing data mechanisms and three proportions of incomplete cases. The missing data mechanisms are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $N \times k$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data and higher values are more likely to be missing. The proportion of incomplete cases $p_{\mathrm inc}$ is set to 25%, 50%, and 75%.

@@

<!-- \begin{table}[] -->
<!-- \begin{tabular}{lllll} -->
<!-- \hline -->
<!-- Missingness &  &                 &  & Early stopping            \\ \cline{1-3} \cline{5-5}  -->
<!-- Mechanism   &  & Proportion      &  & Iteration number          \\ \cline{1-1} \cline{3-3} \cline{5-5}  -->
<!-- MCAR, MAR   &  & 0.25, 0.5, 0.75 &  & 1, 2, 3, ..., 98, 99, 100 \\ \hline -->
<!-- \end{tabular} -->
<!-- \end{table} -->

$$
\begin{aligned}
\hline
&\mathrm{Missingness \ conditions} & & & &\mathrm{Early \ stopping \ conditions} \\
\hline
&\begin{matrix}
\mathrm{MCAR} \\
\mathrm{MAR \ } 
\end{matrix}
\quad
&\times
\quad
&\begin{matrix}
25\% \\
50\% \\
75\%
\end{matrix}
\quad
&\times
\quad
&\begin{matrix}
1 \\
2 \\
... \\
99 \\
100 \\
\end{matrix} \\
&\mathrm{mechansim} &&\mathrm{incomplete \ cases} &&\mathrm{iteration(s)} \\
\hline
\end{aligned}
$$

@@

We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the six missingness conditions.

We impute the missing values in {$y_{\mathrm obs}, \dot y_{\mathrm mis}$} according to the early stopping conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$). All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$). This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\mathrm imp, \ell}$.

Subsequently, we obtain the completed data per imputation, {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].

@@

We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s. Univariate $\theta$s are obtained from $\dot y_{\mathrm imp, \ell}$. The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R]. The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}. We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.

@@

### $Q=\mu$.
The estimated univariate means seem unaffected by early stopping. This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$). The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm inc}\geq.75$ leading to more extreme biases than other conditions. Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}\leq.50$. This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation. Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.

>>>>

---
title: "Non-convergence in iterative imputation"
author: "H. I. Oberman"
abstract: "Iterative imputation has become the de facto standard to accommodate for the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms. We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached. In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value."
# keywords: "missing data, iterative imputation, non-convergence, MICE"
# bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
  fig_width: 6
format: pdf
bibliography: references.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
load("../results/out.RData")
source("../R/evaluate_estimates.R")
performance <- evaluate_est(out)
performance[performance$term == "b0", "term"] <- "Y"
results <- # simulation results split by condition and regression coefficient
performance |> 
  group_by(method, mech, prop, .it, term) |> 
  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) |>
  mutate(mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
         prop = factor(prop, levels = c(0.25, 0.5, 0.75), ordered = TRUE),
         term = factor(term, levels = c("Y", "X1", "X2", "X3"), ordered = TRUE))
```

@@

We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; @R). We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI]. The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{\mathrm{sim}} = 2000$). To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$). We differentiate between six different missingness scenarios that are defined in the data generating mechanism.

@@

In each simulation repetition, we first generate a complete set of $n_{\mathrm{obs}} = 200$ cases, representing person-data in a multiple linear regression problem. The predictor space consists of three multivariately normal random variables,

$$
\begin{pmatrix}
X_1\\
X_2\\
X_3
\end{pmatrix}
\sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
0\\
0\\
0
\end{pmatrix},
\begin{pmatrix}
1     &       &  \\
0.5   & 1     &  \\
0.5  & 0.5   & 1
\end{pmatrix}
\end{bmatrix}.
$$ 
The outcome variable $Y$ is a linear combination of the three predictors, such that for each unit $i = 1, 2,..., n_\mathrm{obs}$,

$$
Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 1)$. This results in a complete dataset of size $n_{\mathrm{obs}} \times p$, with units $i = 1, 2,..., n_\mathrm{obs}$ and variables $j =  Y, X_1, X_2, X_3$). Multivariate normal data are generated using the `mvtnorm` package [@mvtnorm].

Subsequently, the complete are 'amputed' (i.e., made incomplete) according to six missingness conditions. We use a $2 \times 3$ factorial design with two missing data mechanisms and three proportions of incomplete cases, see Table XYZ.

[TODO: make this a box] Missingness mechanisms refer to the probability of being missing for any given entry in a dataset. There are three distinct types of mechanisms, as defined by Rubin (1976). Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), may depend on observed information (MAR; Missing At Random), or may depend on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).

The missing data mechanisms under consideration are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $n_\mathrm{obs} \times p$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data, and higher values are more likely to be missing. The proportion of incomplete cases $\pi_{\mathrm{inc}}$ is set to 25%, 50%, and 75%. 
We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for every missingness condition. The missing data patterns are all possible combinations of multivariate missingness.

[TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?].

@@

We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice]. On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest. Our estimands are the regression coefficients $\boldsymbol{\beta}$,

@@

where $\hat{Y}$ is the predicted value of the outcome. We estimate these coefficients using the `lm()` function [@R] in each imputed dataset, and subsequently pool the estimates across the five imputations using Rubin's rules [function `mice::pool()`; @mice].

<!-- All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$). This results in $m=5$ sets of imputations for each of the conditions, $\dot y_{\mathrm{imp}, \ell}$.  -->

<!-- Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice]. -->

We evaluate the pooled estimates against their true population values using the performance measures bias, confidence interval width, and coverage rate, as recommended by van Buuren (-@buur18 $\S$ 2.5.2). We calculate bias as $\bar{Q} - Q$. CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$. CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

<!-- We evaluate the diagnostic methods to identify non-convergence against these performance measures. $\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$. -->

@@

<!-- We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s. Univariate $\theta$s are obtained from $\dot y_{\mathrm{imp}, \ell}$. The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R]. The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}. We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.  -->

@@

```{r ac_mean}
results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = ac_mean, color = mech, linetype = mech)) +
  geom_hline(yintercept = 0, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()
```

```{r ac_sd}
results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = ac_sd, color = mech, linetype = mech)) +
  geom_hline(yintercept = 0, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()
```

@@

```{r psrf_mean}
results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = psrf_mean, color = mech, linetype = mech)) +
  geom_hline(yintercept = 1, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()
```

```{r psrf_sd}
results |>
  filter(method == "MICE") |> 
  ggplot(aes(x = .it, y = psrf_sd, color = mech, linetype = mech)) +
  geom_hline(yintercept = 1, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ prop) +
  theme_classic()
```

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $y_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot y_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $y_{\mathrm{obs}}$ to create $m$ completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm{inc}}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\mathrm{obs}}$; the missing part of $y$ is referred to as $y_{\mathrm{mis}}$. For each datapoint in $y_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$). The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.

@@

To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays two scenarios from the example. The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations). The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$. This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$. $\bar{Q}$ is then a  biased, invalid estimate of $Q$. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.

@@

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation. The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}. By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.

@@

To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'. Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\mathrm{inc}}$. Early stopping is defined by the number of iterations in the imputation algorithm, $T$.

The missingness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\mathrm{inc}}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. A higher $p_{\mathrm{inc}}$ yields slower convergence---unless there is so little information in $y_{\mathrm{obs}}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that missingness conditions with higher $p_{\mathrm{inc}}$ will result in more signs of non-convergence.

@@

1. We expect that simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.

@@

$$
Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 1)$. This results in a dataset $y$, where $y_{\mathrm{mis}} = \emptyset$. In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).

@@

In each simulation repetition, we *ampute* the complete dataset conform six missingness conditions: two missing data mechanisms and three proportions of incomplete cases. The missing data mechanisms are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $N \times k$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data and higher values are more likely to be missing. The proportion of incomplete cases $p_{\mathrm{inc}}$ is set to 25%, 50%, and 75%.

@@

@@

We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for each of the six missingness conditions.

We impute the missing values in {$y_{\mathrm{obs}}, \dot y_{\mathrm{mis}}$} according to the early stopping conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$). All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$). This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\mathrm{imp}, \ell}$.

Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].

@@

We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s. Univariate $\theta$s are obtained from $\dot y_{\mathrm{imp}, \ell}$. The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R]. The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}. We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.

@@

### $Q=\mu$.
The estimated univariate means seem unaffected by early stopping. This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$). The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions. Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.50$. This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation. Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
